---
title: "Figure 2 - Sysmex analysis"
author: "TP"
date: "2024-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. preperation
## load packages
```{r}
library(variancePartition)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(openxlsx)
library(naniar)
library(rstatix)
library(patchwork)
library(ggsci)
library(RColorBrewer)
library(tidyverse)
library(ggrepel)
```
## colors 
```{r}
# col_timepoint
col_timepoint<- brewer.pal(n = 3, name = "Dark2")
names(col_timepoint)<-c("t0", "t1", "t2")

col_diet_diet <- c("#0564ad", "#b91c1c", "#f1c232")
names(col_diet_diet) <- c("rural_diet", "urban_diet", "MBEGE")
```


# ---------------------------

# 2. load data
## sysmex table
```{r}
sysmex <- read.csv("./241118_sysmex_df.csv", row.names = 1)
head(sysmex)
```

## sample_table
```{r}
sample_table <- read.csv("./241118_sample_table_sysmex.csv", row.names = 1)
head(sample_table)
```





## factor parameters
```{r}
sample_table$diet <- factor(sample_table$diet, levels = c("MBEGE","rural_diet","urban_diet","controls_u", "controls_r"))

sample_table$timepoint <- factor(sample_table$timepoint, levels = c("t0","t1","t2"))
```



```{r}
identical(rownames(sysmex), rownames(sample_table))
```





are all the sysmex IDs present in the sample table
```{r}
all(rownames(sysmex) %in% sample_table$ID)
```


# ===============================


# Samples available for analysis

```{r}
table(sample_table$diet, sample_table$timepoint)
```


# ==============================

### 1. Urban diet











```{r}
order_variables <- c("WBC", "NEUT_A", "IG_A", "IG_P","NEUT_GI","NEUT_RI","MONO_A","RE_MONO_A","RE_Mono_P", "RE_MONO_PM", "RE_LYMP_A","RE_LYMP_P_L", "BASO_P")

```


## *Dream* : Urban diet
### prepare data
### ... info table
```{r}
info_urbandiet_diet <- subset(sample_table, diet == "urban_diet")
rownames(info_urbandiet_diet) <- info_urbandiet_diet$ID

table(info_urbandiet_diet$diet, info_urbandiet_diet$timepoint)
```
### ... sysmex table
```{r}
sysmex_urbandiet_diet_infl <- sysmex[match(rownames(info_urbandiet_diet), rownames(sysmex)),]

vis_miss(sysmex_urbandiet_diet_infl)
```
### ... log transform
```{r}
sysmex_urbandiet_diet_infl_log <- apply(sysmex_urbandiet_diet_infl,2, function(x){
  log2(x + 0.01)
})

sysmex_urbandiet_diet_infl_log <- as.data.frame(sysmex_urbandiet_diet_infl_log)

head(sysmex_urbandiet_diet_infl_log)
```

```{r}
analysis_cols <- c("WBC", "NEUT_A", "IG_A", "IG_P","NEUT_GI","NEUT_RI","MONO_A","RE_MONO_A","RE_Mono_P", "RE_MONO_PM", "RE_LYMP_A","RE_LYMP_P_L", "BASO_P")

sysmex_urbandiet_diet_infl_log <- sysmex_urbandiet_diet_infl_log[,analysis_cols]

vis_miss(sysmex_urbandiet_diet_infl_log)
```

```{r}
identical(rownames(sysmex_urbandiet_diet_infl_log), rownames(info_urbandiet_diet))
```

```{r}
## factor timepoints
info_urbandiet_diet$timepoint <- factor(info_urbandiet_diet$timepoint, levels = c("t0","t1","t2"))

unique(info_urbandiet_diet$timepoint )
```
```{r}
colnames(info_urbandiet_diet)
```

```{r}
L <- makeContrastsDream(~ 0 + timepoint + Age + BMI_t0 + activity_sum_rank_scaled + (1|PID) , info_urbandiet_diet,
  contrasts = c(
    t0_vs_t1 = "timepointt0 - timepointt1",
    t0_vs_t2 = "timepointt0 - timepointt2",
    t1_vs_t2 = "timepointt1 - timepointt2"
  )
)

# Visualize contrast matrix
plotContrasts(L)
```
```{r}
fitmm_urbandiet_diet <- dream(t(sysmex_urbandiet_diet_infl_log), ~ 0 + timepoint + Age + BMI_t0 + activity_sum_rank_scaled + (1|PID)  , info_urbandiet_diet, L, ddf = "Kenward-Roger", useWeights = FALSE)

fitmm_urbandiet_diet <- eBayes(fitmm_urbandiet_diet)
```

```{r}
# Examine design matrix
head(fitmm_urbandiet_diet$design, 3)
```

```{r}
colnames(fitmm_urbandiet_diet)

```

```{r}
## create a list of results
dream_urbandiet_diet <- data.frame() # create empty dataframe for storage

for(i in  colnames(L)){
  print(i)
  
  tmpi <- variancePartition::topTable(fitmm_urbandiet_diet, coef = i, number = Inf, adjust.method = "BH", sort.by = "p", lfc = 0)
  
  tmpi$comparison <- i
  
  tmpi$variable <- rownames(tmpi)
  
  tmpi <- as.data.frame(tmpi %>% add_significance(p.col = "adj.P.Val"))
  
  dream_urbandiet_diet <- rbind(dream_urbandiet_diet, tmpi)
  
}

head(dream_urbandiet_diet)
```
#### ... add paired log2FC
calculate the log2FC2 values for later timepoint vs early point for visualization
```{r}
identical(rownames(info_urbandiet_diet), rownames(sysmex_urbandiet_diet_infl_log))
```

```{r}
pairedFC_urbandiet_diet <- data.frame()

for ( i in colnames(L)){
  print(i)
  
  ctr_condition = str_split_fixed(i, "_vs_",2)[,1]
  test_condition = str_split_fixed(i, "_vs_",2)[,2]
  
  print(ctr_condition)
  print(test_condition)
  
  ctrids <- rownames(subset(info_urbandiet_diet, as.character(timepoint) == ctr_condition))
  testids <- rownames(subset(info_urbandiet_diet, as.character(timepoint) == test_condition))

  tmpfc <- fcros::fcrosFCmat(t(sysmex_urbandiet_diet_infl_log), ctrids, testids, log2.opt = 0)

  tmp_pairedFC <- data.frame(variable = tmpfc$idnames, 
                                       FC = as.numeric(tmpfc$FC), 
                                       FC2 = as.numeric(tmpfc$FC2))

  tmp_pairedFC$log2FC <- log2(tmp_pairedFC$FC)
  tmp_pairedFC$log2FC2 <- log2(tmp_pairedFC$FC2)

  tmp_pairedFC$direction <- ifelse(tmp_pairedFC$log2FC2>0,"up",  "down")

  tmp_pairedFC$comparison <- i
  
  pairedFC_urbandiet_diet <- rbind(pairedFC_urbandiet_diet, tmp_pairedFC)
  
}

head(pairedFC_urbandiet_diet)
```

```{r}
### add paired FC
calculate the log2FC2 values for later timepoint vs early point for visualization
pairedFC_urbandiet_diet$merged <- paste0(pairedFC_urbandiet_diet$variable,"_", pairedFC_urbandiet_diet$comparison)

dream_urbandiet_diet$merged <- paste0(dream_urbandiet_diet$variable,"_", dream_urbandiet_diet$comparison)

cols_2_add <- c("FC","FC2","log2FC","log2FC2","direction","merged")

dream_urbandiet_diet <- merge(dream_urbandiet_diet, pairedFC_urbandiet_diet[,cols_2_add], by = "merged")

dream_urbandiet_diet$regulation <- ifelse(dream_urbandiet_diet$adj.P.Val.signif != "ns", 
                                         dream_urbandiet_diet$direction, "ns")

rownames(dream_urbandiet_diet) <- dream_urbandiet_diet$merged

head(dream_urbandiet_diet)
```

```{r}
list_sig_dream_urbandiet_diet <- list()

for(i in unique(dream_urbandiet_diet$comparison)){
  
  print(i)
  
  list_sig_dream_urbandiet_diet[[i]] <- split(
    subset(dream_urbandiet_diet, adj.P.Val.signif != "ns" & comparison == i)$variable,
     subset(dream_urbandiet_diet, adj.P.Val.signif != "ns" & comparison == i)$regulation
    )
  
}

str(list_sig_dream_urbandiet_diet)
```
```{r}
list_sig_dream_urbandiet_diet
```


# -----------------------

## *Dream* : Urban diet controls
### prepare data
### ... info table
```{r}
info_urbandiet_ctr <- subset(sample_table, diet == "controls_r")
rownames(info_urbandiet_ctr) <- info_urbandiet_ctr$ID

table(info_urbandiet_ctr$diet, info_urbandiet_ctr$timepoint)
```
### ... sysmex table
```{r}
sysmex_urbandiet_ctr_infl <- sysmex[match(rownames(info_urbandiet_ctr), rownames(sysmex)),]

vis_miss(sysmex_urbandiet_ctr_infl)
```
### ... log transform
```{r}
sysmex_urbandiet_ctr_infl_log <- apply(sysmex_urbandiet_ctr_infl, 2, function(x){
  log2(x + 0.01)
})

sysmex_urbandiet_ctr_infl_log <- as.data.frame(sysmex_urbandiet_ctr_infl_log)

head(sysmex_urbandiet_ctr_infl_log)
```


```{r}
analysis_cols <- c("WBC", "NEUT_A", "IG_A", "IG_P","NEUT_GI","NEUT_RI","MONO_A","RE_MONO_A","RE_Mono_P", "RE_MONO_PM", "RE_LYMP_A","RE_LYMP_P_L", "BASO_P")

sysmex_urbandiet_ctr_infl_log <- sysmex_urbandiet_ctr_infl_log[,analysis_cols]

vis_miss(sysmex_urbandiet_ctr_infl_log)
```

```{r}
identical(rownames(sysmex_urbandiet_ctr_infl_log), rownames(info_urbandiet_ctr))
```

```{r}
## factor timepoints
info_urbandiet_ctr$timepoint <- factor(info_urbandiet_ctr$timepoint, levels = c("t0","t1","t2"))

unique(info_urbandiet_ctr$timepoint )
```
```{r}
colnames(info_urbandiet_ctr)
```

```{r}
L <- makeContrastsDream(~ 0 + timepoint + Age + BMI_t0 + activity_sum_rank_scaled + (1|PID) , info_urbandiet_ctr,
  contrasts = c(
    t0_vs_t1 = "timepointt0 - timepointt1",
    t0_vs_t2 = "timepointt0 - timepointt2",
    t1_vs_t2 = "timepointt1 - timepointt2"
  )
)

# Visualize contrast matrix
plotContrasts(L)
```
```{r}
fitmm_urbandiet_ctr <- dream(t(sysmex_urbandiet_ctr_infl_log), ~ 0 + timepoint + Age + BMI_t0 + activity_sum_rank_scaled + (1|PID)  , info_urbandiet_ctr, L, ddf = "Kenward-Roger", useWeights = FALSE)

fitmm_urbandiet_ctr <- eBayes(fitmm_urbandiet_ctr)
```

```{r}
# Examine design matrix
head(fitmm_urbandiet_ctr$design, 3)
```

```{r}
colnames(fitmm_urbandiet_ctr)

```

```{r}
## create a list of results
dream_urbandiet_ctr <- data.frame() # create empty dataframe for storage

for(i in  colnames(L)){
  print(i)
  
  tmpi <- variancePartition::topTable(fitmm_urbandiet_ctr, coef = i, number = Inf, adjust.method = "BH", sort.by = "p", lfc = 0)
  
  tmpi$comparison <- i
  
  tmpi$variable <- rownames(tmpi)
  
  tmpi <- as.data.frame(tmpi %>% add_significance(p.col = "adj.P.Val"))
  
  dream_urbandiet_ctr <- rbind(dream_urbandiet_ctr, tmpi)
  
}

head(dream_urbandiet_ctr)
```
#### ... add paired log2FC
calculate the log2FC2 values for later timepoint vs early point for visualization
```{r}
identical(rownames(info_urbandiet_ctr), rownames(sysmex_urbandiet_ctr_infl_log))
```

```{r}
pairedFC_urbandiet_ctr <- data.frame()

for ( i in colnames(L)){
  print(i)
  
  ctr_condition = str_split_fixed(i, "_vs_",2)[,1]
  test_condition = str_split_fixed(i, "_vs_",2)[,2]
  
  print(ctr_condition)
  print(test_condition)
  
  ctrids <- rownames(subset(info_urbandiet_ctr, as.character(timepoint) == ctr_condition))
  testids <- rownames(subset(info_urbandiet_ctr, as.character(timepoint) == test_condition))

  tmpfc <- fcros::fcrosFCmat(t(sysmex_urbandiet_ctr_infl_log), ctrids, testids, log2.opt = 0)

  tmp_pairedFC <- data.frame(variable = tmpfc$idnames, 
                                       FC = as.numeric(tmpfc$FC), 
                                       FC2 = as.numeric(tmpfc$FC2))

  tmp_pairedFC$log2FC <- log2(tmp_pairedFC$FC)
  tmp_pairedFC$log2FC2 <- log2(tmp_pairedFC$FC2)

  tmp_pairedFC$direction <- ifelse(tmp_pairedFC$log2FC2>0,"up",  "down")

  tmp_pairedFC$comparison <- i
  
  pairedFC_urbandiet_ctr <- rbind(pairedFC_urbandiet_ctr, tmp_pairedFC)
  
}

head(pairedFC_urbandiet_ctr)
```

```{r}
### add paired FC
## calculate the log2FC2 values for later timepoint vs early point for visualization
pairedFC_urbandiet_ctr$merged <- paste0(pairedFC_urbandiet_ctr$variable,"_", pairedFC_urbandiet_ctr$comparison)

dream_urbandiet_ctr$merged <- paste0(dream_urbandiet_ctr$variable,"_", dream_urbandiet_ctr$comparison)

cols_2_add <- c("FC","FC2","log2FC","log2FC2","direction","merged")

dream_urbandiet_ctr <- merge(dream_urbandiet_ctr, pairedFC_urbandiet_ctr[,cols_2_add], by = "merged")

dream_urbandiet_ctr$regulation <- ifelse(dream_urbandiet_ctr$adj.P.Val.signif != "ns", 
                                         dream_urbandiet_ctr$direction, "ns")

rownames(dream_urbandiet_ctr) <- dream_urbandiet_ctr$merged

head(dream_urbandiet_ctr)
```

```{r}
list_sig_dream_urbandiet_ctr <- list()

for(i in unique(dream_urbandiet_ctr$comparison)){
  
  print(i)
  
  list_sig_dream_urbandiet_ctr[[i]] <- split(
    subset(dream_urbandiet_ctr, adj.P.Val.signif != "ns" & comparison == i)$variable,
     subset(dream_urbandiet_ctr, adj.P.Val.signif != "ns" & comparison == i)$regulation
    )
  
}

str(list_sig_dream_urbandiet_ctr)
```
```{r}
list_sig_dream_urbandiet_ctr
```


# ==============================

### 2. Rural diet

## *Dream* : Rural diet
### prepare data
### ... info table
```{r}
info_ruraldiet_diet <- subset(sample_table, diet == "rural_diet")
rownames(info_ruraldiet_diet) <- info_ruraldiet_diet$ID

table(info_ruraldiet_diet$diet, info_ruraldiet_diet$timepoint)
```

### ... sysmex table
```{r}
sysmex_ruraldiet_diet_infl <- sysmex[match(rownames(info_ruraldiet_diet), rownames(sysmex)),]

vis_miss(sysmex_ruraldiet_diet_infl)
```

### ... log transform
```{r}
sysmex_ruraldiet_diet_infl_log <- apply(sysmex_ruraldiet_diet_infl, 2, function(x){
  log2(x + 0.01)
})

sysmex_ruraldiet_diet_infl_log <- as.data.frame(sysmex_ruraldiet_diet_infl_log)

head(sysmex_ruraldiet_diet_infl_log)
```


```{r}
analysis_cols <- c("WBC", "NEUT_A", "IG_A", "IG_P","NEUT_GI","NEUT_RI","MONO_A","RE_MONO_A","RE_Mono_P", "RE_MONO_PM", "RE_LYMP_A","RE_LYMP_P_L", "BASO_P")

sysmex_ruraldiet_diet_infl_log <- sysmex_ruraldiet_diet_infl_log[,analysis_cols]

vis_miss(sysmex_ruraldiet_diet_infl_log)
```

```{r}
identical(rownames(sysmex_ruraldiet_diet_infl_log), rownames(info_ruraldiet_diet))
```

```{r}
## factor timepoints
info_ruraldiet_diet$timepoint <- factor(info_ruraldiet_diet$timepoint, levels = c("t0","t1","t2"))

unique(info_ruraldiet_diet$timepoint )
```
```{r}
colnames(info_ruraldiet_diet)
```

```{r}
L <- makeContrastsDream(~ 0 + timepoint + Age + BMI_t0 + activity_sum_rank_scaled + (1|PID) , info_ruraldiet_diet,
  contrasts = c(
    t0_vs_t1 = "timepointt0 - timepointt1",
    t0_vs_t2 = "timepointt0 - timepointt2",
    t1_vs_t2 = "timepointt1 - timepointt2"
  )
)

# Visualize contrast matrix
plotContrasts(L)
```
```{r}
fitmm_ruraldiet_diet <- dream(t(sysmex_ruraldiet_diet_infl_log), ~ 0 + timepoint + Age + BMI_t0 + activity_sum_rank_scaled + (1|PID)  , info_ruraldiet_diet, L, ddf = "Kenward-Roger", useWeights = FALSE)

fitmm_ruraldiet_diet <- eBayes(fitmm_ruraldiet_diet)
```

```{r}
# Examine design matrix
head(fitmm_ruraldiet_diet$design, 3)
```

```{r}
colnames(fitmm_ruraldiet_diet)

```

```{r}
## create a list of results
dream_ruraldiet_diet <- data.frame() # create empty dataframe for storage

for(i in  colnames(L)){
  print(i)
  
  tmpi <- variancePartition::topTable(fitmm_ruraldiet_diet, coef = i, number = Inf, adjust.method = "BH", sort.by = "p", lfc = 0)
  
  tmpi$comparison <- i
  
  tmpi$variable <- rownames(tmpi)
  
  tmpi <- as.data.frame(tmpi %>% add_significance(p.col = "adj.P.Val"))
  
  dream_ruraldiet_diet <- rbind(dream_ruraldiet_diet, tmpi)
  
}

head(dream_ruraldiet_diet)
```
#### ... add paired log2FC
calculate the log2FC2 values for later timepoint vs early point for visualization
```{r}
identical(rownames(info_ruraldiet_diet), rownames(sysmex_ruraldiet_diet_infl_log))
```

```{r}
pairedFC_ruraldiet_diet <- data.frame()

for ( i in colnames(L)){
  print(i)
  
  ctr_condition = str_split_fixed(i, "_vs_",2)[,1]
  test_condition = str_split_fixed(i, "_vs_",2)[,2]
  
  print(ctr_condition)
  print(test_condition)
  
  ctrids <- rownames(subset(info_ruraldiet_diet, as.character(timepoint) == ctr_condition))
  testids <- rownames(subset(info_ruraldiet_diet, as.character(timepoint) == test_condition))

  tmpfc <- fcros::fcrosFCmat(t(sysmex_ruraldiet_diet_infl_log), ctrids, testids, log2.opt = 0)

  tmp_pairedFC <- data.frame(variable = tmpfc$idnames, 
                                       FC = as.numeric(tmpfc$FC), 
                                       FC2 = as.numeric(tmpfc$FC2))

  tmp_pairedFC$log2FC <- log2(tmp_pairedFC$FC)
  tmp_pairedFC$log2FC2 <- log2(tmp_pairedFC$FC2)

  tmp_pairedFC$direction <- ifelse(tmp_pairedFC$log2FC2>0,"up",  "down")

  tmp_pairedFC$comparison <- i
  
  pairedFC_ruraldiet_diet <- rbind(pairedFC_ruraldiet_diet, tmp_pairedFC)
  
}

head(pairedFC_ruraldiet_diet)
```

```{r}
### add paired FC
pairedFC_ruraldiet_diet$merged <- paste0(pairedFC_ruraldiet_diet$variable,"_", pairedFC_ruraldiet_diet$comparison)

dream_ruraldiet_diet$merged <- paste0(dream_ruraldiet_diet$variable,"_", dream_ruraldiet_diet$comparison)

cols_2_add <- c("FC","FC2","log2FC","log2FC2","direction","merged")

dream_ruraldiet_diet <- merge(dream_ruraldiet_diet, pairedFC_ruraldiet_diet[,cols_2_add], by = "merged")

dream_ruraldiet_diet$regulation <- ifelse(dream_ruraldiet_diet$adj.P.Val.signif != "ns", 
                                         dream_ruraldiet_diet$direction, "ns")

rownames(dream_ruraldiet_diet) <- dream_ruraldiet_diet$merged

head(dream_ruraldiet_diet)
```

```{r}
list_sig_dream_ruraldiet_diet <- list()

for(i in unique(dream_ruraldiet_diet$comparison)){
  
  print(i)
  
  list_sig_dream_ruraldiet_diet[[i]] <- split(
    subset(dream_ruraldiet_diet, adj.P.Val.signif != "ns" & comparison == i)$variable,
     subset(dream_ruraldiet_diet, adj.P.Val.signif != "ns" & comparison == i)$regulation
    )
  
}

str(list_sig_dream_ruraldiet_diet)
```
```{r}
list_sig_dream_ruraldiet_diet
```


# ---------------------------------------

### 2.2. urban controls
## *Dream* : urban controls
### prepare data
### ... info table
```{r}
info_ruraldiet_contol <- subset(sample_table, diet == "controls_u")
rownames(info_ruraldiet_contol) <- info_ruraldiet_contol$ID

table(info_ruraldiet_contol$diet, info_ruraldiet_contol$timepoint)
```

### ... sysmex table
```{r}
sysmex_ruraldiet_ctr_infl <- sysmex[match(rownames(info_ruraldiet_contol), rownames(sysmex)),]

vis_miss(sysmex_ruraldiet_ctr_infl)
```
### ... log transform
```{r}
sysmex_ruraldiet_ctr_infl_log <- apply(sysmex_ruraldiet_ctr_infl, 2, function(x){
  log2(x + 0.01)
})

sysmex_ruraldiet_ctr_infl_log <- as.data.frame(sysmex_ruraldiet_ctr_infl_log)

head(sysmex_ruraldiet_ctr_infl_log)
```


```{r}
analysis_cols <- c("WBC", "NEUT_A", "IG_A", "IG_P","NEUT_GI","NEUT_RI","MONO_A","RE_MONO_A","RE_Mono_P", "RE_MONO_PM", "RE_LYMP_A","RE_LYMP_P_L", "BASO_P")

sysmex_ruraldiet_ctr_infl_log <- sysmex_ruraldiet_ctr_infl_log[,analysis_cols]

vis_miss(sysmex_ruraldiet_ctr_infl_log)
```

```{r}
identical(rownames(sysmex_ruraldiet_ctr_infl_log), rownames(info_ruraldiet_contol))
```

```{r}
## factor timepoints
info_ruraldiet_contol$timepoint <- factor(info_ruraldiet_contol$timepoint, levels = c("t0","t1","t2"))

unique(info_ruraldiet_contol$timepoint )
```
```{r}
colnames(info_ruraldiet_contol)
```

```{r}
L <- makeContrastsDream(~ 0 + timepoint + Age + BMI_t0 + activity_sum_rank_scaled + (1|PID) , info_ruraldiet_contol,
  contrasts = c(
    t0_vs_t1 = "timepointt0 - timepointt1",
    t0_vs_t2 = "timepointt0 - timepointt2",
    t1_vs_t2 = "timepointt1 - timepointt2"
  )
)

# Visualize contrast matrix
plotContrasts(L)
```
```{r}
fitmm_ruraldiet_ctr <- dream(t(sysmex_ruraldiet_ctr_infl_log), ~ 0 + timepoint + Age + BMI_t0 + activity_sum_rank_scaled + (1|PID)  , info_ruraldiet_contol, L, ddf = "Kenward-Roger", useWeights = FALSE)

fitmm_ruraldiet_ctr <- eBayes(fitmm_ruraldiet_ctr)
```

```{r}
# Examine design matrix
head(fitmm_ruraldiet_ctr$design, 3)
```

```{r}
colnames(fitmm_ruraldiet_ctr)

```

```{r}
## create a list of results
dream_ruraldiet_ctr <- data.frame() # create empty dataframe for storage

for(i in  colnames(L)){
  print(i)
  
  tmpi <- variancePartition::topTable(fitmm_ruraldiet_ctr, coef = i, number = Inf, adjust.method = "BH", sort.by = "p", lfc = 0)
  
  tmpi$comparison <- i
  
  tmpi$variable <- rownames(tmpi)
  
  tmpi <- as.data.frame(tmpi %>% add_significance(p.col = "adj.P.Val"))
  
  dream_ruraldiet_ctr <- rbind(dream_ruraldiet_ctr, tmpi)
  
}

head(dream_ruraldiet_ctr)
```
#### ... add paired log2FC
calculate the log2FC2 values for later timepoint vs early point for visualization
```{r}
identical(rownames(info_ruraldiet_contol), rownames(sysmex_ruraldiet_ctr_infl_log))
```

```{r}
pairedFC_ruraldiet_ctr <- data.frame()

for ( i in colnames(L)){
  print(i)
  
  ctr_condition = str_split_fixed(i, "_vs_",2)[,1]
  test_condition = str_split_fixed(i, "_vs_",2)[,2]
  
  print(ctr_condition)
  print(test_condition)
  
  ctrids <- rownames(subset(info_ruraldiet_contol, as.character(timepoint) == ctr_condition))
  testids <- rownames(subset(info_ruraldiet_contol, as.character(timepoint) == test_condition))

  tmpfc <- fcros::fcrosFCmat(t(sysmex_ruraldiet_ctr_infl_log), ctrids, testids, log2.opt = 0)

  tmp_pairedFC <- data.frame(variable = tmpfc$idnames, 
                                       FC = as.numeric(tmpfc$FC), 
                                       FC2 = as.numeric(tmpfc$FC2))

  tmp_pairedFC$log2FC <- log2(tmp_pairedFC$FC)
  tmp_pairedFC$log2FC2 <- log2(tmp_pairedFC$FC2)

  tmp_pairedFC$direction <- ifelse(tmp_pairedFC$log2FC2>0,"up",  "down")

  tmp_pairedFC$comparison <- i
  
  pairedFC_ruraldiet_ctr <- rbind(pairedFC_ruraldiet_ctr, tmp_pairedFC)
  
}

head(pairedFC_ruraldiet_ctr)
```

```{r}
### add paired FC
pairedFC_ruraldiet_ctr$merged <- paste0(pairedFC_ruraldiet_ctr$variable,"_", pairedFC_ruraldiet_ctr$comparison)

dream_ruraldiet_ctr$merged <- paste0(dream_ruraldiet_ctr$variable,"_", dream_ruraldiet_ctr$comparison)

cols_2_add <- c("FC","FC2","log2FC","log2FC2","direction","merged")

dream_ruraldiet_ctr <- merge(dream_ruraldiet_ctr, pairedFC_ruraldiet_ctr[,cols_2_add], by = "merged")

dream_ruraldiet_ctr$regulation <- ifelse(dream_ruraldiet_ctr$adj.P.Val.signif != "ns", 
                                         dream_ruraldiet_ctr$direction, "ns")

rownames(dream_ruraldiet_ctr) <- dream_ruraldiet_ctr$merged

head(dream_ruraldiet_ctr)
```

```{r}
list_sig_dream_ruraldiet_ctr <- list()

for(i in unique(dream_ruraldiet_ctr$comparison)){
  
  print(i)
  
  list_sig_dream_ruraldiet_ctr[[i]] <- split(
    subset(dream_ruraldiet_ctr, adj.P.Val.signif != "ns" & comparison == i)$variable,
     subset(dream_ruraldiet_ctr, adj.P.Val.signif != "ns" & comparison == i)$regulation
    )
  
}

str(list_sig_dream_ruraldiet_ctr)
```
```{r}
list_sig_dream_ruraldiet_ctr
```
# ====================================
### 3. Mbege

## *Dream* : MBEGE
### prepare data
### ... info table
```{r}
info_mbege_diet <- subset(sample_table, diet == "MBEGE")
rownames(info_mbege_diet) <- info_mbege_diet$ID

table(info_mbege_diet$diet, info_mbege_diet$timepoint)
```

### ... sysmex table
```{r}
sysmex_mbege_diet_infl <- sysmex[match(rownames(info_mbege_diet), rownames(sysmex)),]

vis_miss(sysmex_mbege_diet_infl)
```
```{r}
sysmex_mbege_diet_infl_log <- apply(sysmex_mbege_diet_infl, 2, function(x){
  log2(x + 0.01)
})

sysmex_mbege_diet_infl_log <- as.data.frame(sysmex_mbege_diet_infl_log)

head(sysmex_mbege_diet_infl_log)
```


```{r}
analysis_cols <- c("WBC", "NEUT_A", "IG_A", "IG_P","NEUT_GI","NEUT_RI","MONO_A","RE_MONO_A","RE_Mono_P", "RE_MONO_PM", "RE_LYMP_A","RE_LYMP_P_L", "BASO_P")

sysmex_mbege_diet_infl_log <- sysmex_mbege_diet_infl_log[,analysis_cols]

vis_miss(sysmex_mbege_diet_infl_log)
```

```{r}
identical(rownames(sysmex_mbege_diet_infl_log), rownames(info_mbege_diet))
```

```{r}
## factor timepoints
info_mbege_diet$timepoint <- factor(info_mbege_diet$timepoint, levels = c("t0","t1","t2"))

unique(info_mbege_diet$timepoint )
```
```{r}
colnames(info_mbege_diet)
```

```{r}
L <- makeContrastsDream(~ 0 + timepoint + Age + BMI_t0 + activity_sum_rank_scaled + (1|PID) , info_mbege_diet,
  contrasts = c(
    t0_vs_t1 = "timepointt0 - timepointt1",
    t0_vs_t2 = "timepointt0 - timepointt2",
    t1_vs_t2 = "timepointt1 - timepointt2"
  )
)

# Visualize contrast matrix
plotContrasts(L)
```
```{r}
fitmm_mbege_diet <- dream(t(sysmex_mbege_diet_infl_log), ~ 0 + timepoint + Age + BMI_t0 + activity_sum_rank_scaled + (1|PID)  , info_mbege_diet, L, ddf = "Kenward-Roger", useWeights = FALSE)

fitmm_mbege_diet <- eBayes(fitmm_mbege_diet)
```

```{r}
# Examine design matrix
head(fitmm_mbege_diet$design, 3)
```

```{r}
colnames(fitmm_mbege_diet)

```

```{r}
## create a list of results
dream_mbege_diet <- data.frame() # create empty dataframe for storage

for(i in  colnames(L)){
  print(i)
  
  tmpi <- variancePartition::topTable(fitmm_mbege_diet, coef = i, number = Inf, adjust.method = "BH", sort.by = "p", lfc = 0)
  
  tmpi$comparison <- i
  
  tmpi$variable <- rownames(tmpi)
  
  tmpi <- as.data.frame(tmpi %>% add_significance(p.col = "adj.P.Val"))
  
  dream_mbege_diet <- rbind(dream_mbege_diet, tmpi)
  
}

head(dream_mbege_diet)
```
#### ... add paired log2FC
```{r}
identical(rownames(info_mbege_diet), rownames(sysmex_mbege_diet_infl_log))
```

```{r}
pairedFC_mbege_diet <- data.frame()

for ( i in colnames(L)){
  print(i)
  
  ctr_condition = str_split_fixed(i, "_vs_",2)[,1]
  test_condition = str_split_fixed(i, "_vs_",2)[,2]
  
  print(ctr_condition)
  print(test_condition)
  
  ctrids <- rownames(subset(info_mbege_diet, as.character(timepoint) == ctr_condition))
  testids <- rownames(subset(info_mbege_diet, as.character(timepoint) == test_condition))

  tmpfc <- fcros::fcrosFCmat(t(sysmex_mbege_diet_infl_log), ctrids, testids, log2.opt = 0)

  tmp_pairedFC <- data.frame(variable = tmpfc$idnames, 
                                       FC = as.numeric(tmpfc$FC), 
                                       FC2 = as.numeric(tmpfc$FC2))

  tmp_pairedFC$log2FC <- log2(tmp_pairedFC$FC)
  tmp_pairedFC$log2FC2 <- log2(tmp_pairedFC$FC2)

  tmp_pairedFC$direction <- ifelse(tmp_pairedFC$log2FC2>0,"up",  "down")

  tmp_pairedFC$comparison <- i
  
  pairedFC_mbege_diet <- rbind(pairedFC_mbege_diet, tmp_pairedFC)
  
}

head(pairedFC_mbege_diet)
```

```{r}
### add paired FC
pairedFC_mbege_diet$merged <- paste0(pairedFC_mbege_diet$variable,"_", pairedFC_mbege_diet$comparison)

dream_mbege_diet$merged <- paste0(dream_mbege_diet$variable,"_", dream_mbege_diet$comparison)

cols_2_add <- c("FC","FC2","log2FC","log2FC2","direction","merged")

dream_mbege_diet <- merge(dream_mbege_diet, pairedFC_mbege_diet[,cols_2_add], by = "merged")

dream_mbege_diet$regulation <- ifelse(dream_mbege_diet$adj.P.Val.signif != "ns", 
                                         dream_mbege_diet$direction, "ns")

rownames(dream_mbege_diet) <- dream_mbege_diet$merged

head(dream_mbege_diet)
```

```{r}
list_sig_dream_mbege_diet <- list()

for(i in unique(dream_mbege_diet$comparison)){
  
  print(i)
  
  list_sig_dream_mbege_diet[[i]] <- split(
    subset(dream_mbege_diet, adj.P.Val.signif != "ns" & comparison == i)$variable,
     subset(dream_mbege_diet, adj.P.Val.signif != "ns" & comparison == i)$regulation
    )
  
}

str(list_sig_dream_mbege_diet)
```
```{r}
list_sig_dream_mbege_diet
```



# ==========================

# Summary results file
### summary table
```{r}
summary_res_dream <- list(urban_diet = dream_urbandiet_diet, 
                          rural_diet = dream_ruraldiet_diet, 
                          MBEGE = dream_mbege_diet, 
                          controls_u = dream_ruraldiet_ctr, 
                          controls_r = dream_urbandiet_ctr)

summary_res_dream <- do.call(rbind, Map(cbind, diet = names(summary_res_dream), summary_res_dream))

head(summary_res_dream)
```
```{r}
plot_order_vars <- c("WBC", "NEUT_A", "IG_A", "IG_P", "NEUT_GI", "NEUT_RI", "BASO_P","MONO_A","RE_MONO_A","RE_Mono_P","RE_MONO_PM","RE_LYMP_A","RE_LYMP_P_L" )
```


```{r}
summary_res_dream$variable <- factor(summary_res_dream$variable, levels = rev(plot_order_vars))

summary_res_dream$comparison <- factor(summary_res_dream$comparison, levels = c("t0_vs_t1","t1_vs_t2","t0_vs_t2"))

summary_res_dream$diet <- factor(summary_res_dream$diet, levels = c("urban_diet","rural_diet","MBEGE","controls_r","controls_u"))
```


### plot
```{r}
plotdf <- subset(summary_res_dream, diet %in% c("urban_diet","rural_diet","MBEGE"))
plotdf$label <- ifelse(plotdf$adj.P.Val.signif != "ns", plotdf$adj.P.Val.signif, NA)
myshapes <- c("up" = 24, "down" = 25,"ns" = NA)
max_val <- round(max(plotdf$log2FC2),1)*1.2

summary_plot <-  ggplot(plotdf, aes(x= comparison, y= variable, fill= log2FC2))+
  geom_point(size = 5, aes(shape = regulation))+
  scale_shape_manual(values = myshapes)+
  facet_wrap(.~diet, nrow = 1)+
  geom_text(aes(label = label), size = 6, fontface = "bold")+
  theme_bw()+
  scale_fill_gradientn(colours = c("blue","white","red"), limits = c(-max_val, max_val),
                       na.value = "white")+
  theme(axis.text.x = element_text(angle = 90), 
        axis.text.y = element_text())

summary_plot
```

```{r}
# ## save sysmex table
# write.csv(sample_table,"./241118_sample_table_sysmex.csv")
# ## save sample table table
# write.csv(sysmex[,analysis_cols], "./241118_sysmex_df.csv")
```



# =================================================

# 4. Session info

```{r}
sessionInfo()
```

